---
title: "functions"
output: html_document
date: "2026-05-21"
---

These are the functions that I have used in running the code. Some codes might not be necessary. The code I have provided should include the but please run this code first. 


Function to resample raster files and save them into stack rasters in the same directory
```{r}
function_6_to_5arcmin_stack <- function(in_directory, resample_method) {
  library(terra)
  library(tools)
  
  in_dir <- in_directory
  out_dir <- paste0(in_directory, "/5arcmin")
  
  if (!dir.exists(out_dir)) {
    dir.create(out_dir)
  }
  
  
  original_files <- list.files(
    path = in_dir,
    pattern = ".tif$",
    full.names = TRUE
  )
  
  r_ref = rast(ncol=4320, nrow = 2160, crs="EPSG:4326")
  
  for (file_path in original_files){
    r_original = rast(file_path)
    
    r_resampled = resample(r_original, r_ref, method = resample_method)
    
    base_name <- file_path_sans_ext(basename(file_path))
    
    out_filename <- paste0(base_name, "_5arcmin.tif")
    out_path <- file.path(out_dir, out_filename)
    
    # Save the resampled raster
    writeRaster(r_resampled, out_path, overwrite = TRUE)
    
    # Print a message to track progress
    message("Resampled and saved: ", out_filename)
  }
  message("Resampling complete for all files.")
  
  files_to_stack <- list.files(
    path = out_dir,
    pattern = "_5arcmin\\.tif$",
    full.names = TRUE
  ) |> sort()
  
  years <- as.numeric(gsub(".*?(\\d{4}).*", "\\1", basename(files_to_stack)))
  
  raster_stack <- rast(files_to_stack)
  
  new_names <- paste0("y", years)
  names(raster_stack) <- new_names
  
  # --- Automate the output filename ---
  # 1. Get the first filename as a template
  first_file <- basename(files_to_stack[1])
  
  # 2. Remove the year part (e.g., "_1992")
  name_sans_year <- gsub("_\\d{4}", "", first_file)
  
  # 3. Remove the file extension
  base_name <- file_path_sans_ext(name_sans_year)
  
  # 4. Create the final stack filename
  stack_filename <- paste0(base_name, "_stack.tif")
  
  out_path <- file.path(out_dir, stack_filename)
  writeRaster(raster_stack, out_path, overwrite = TRUE)
  
  message("Raster stack saved to: ", out_path)
}
```



Function to stack all the raster files in the folder and save the names according to the year
```{r}
function_raster_stack <- function(in_directory) {
  library(terra)
  library(tools)
  
  files_to_stack <- list.files(
    path = in_directory,
    pattern = "\\.tif$",
    full.names = TRUE
  ) |> sort()
  
  files_to_stack <- files_to_stack[
    grepl("(19[0-9]{2}|20[0-9]{2})", basename(files_to_stack)) &
      !grepl("stack", basename(files_to_stack), ignore.case = TRUE)
  ]
  
  years <- as.numeric(sub(".*?(\\d{4}).*", "\\1", basename(files_to_stack)))
  
  # Keep only valid year files
  keep <- !is.na(years)
  files_to_stack <- files_to_stack[keep]
  years <- years[keep]
  
  raster_stack <- rast(files_to_stack)
  
  # Layer names: y1992 ... y2020
  names(raster_stack) <- paste0("y", years)
  
  # Build output name from first file, removing year
  first_file <- basename(files_to_stack[1])
  name_sans_year <- gsub("_?\\d{4}", "", first_file)
  base_name <- file_path_sans_ext(name_sans_year)
  stack_filename <- paste0(base_name, "_stack.tif")
  
  out_path <- file.path(in_directory, stack_filename)
  writeRaster(raster_stack, out_path, overwrite = TRUE)
  
  message("Raster stack saved to: ", out_path)
  
  return(raster_stack)
}
```


This is used for masking and clipping the files. This function is written by Reeti Nykänen (reeti.nykanen@aalto.fi)
```{r}
# ------------------------------------------------------------------
# mask + clip (60S)  -> write *_MASKED.tif
# ------------------------------------------------------------------
mask_one <- function(infile, out_file, mask_r, clip_ext) {
  r <- rast(infile)
  
  # 1) crop the variable first (so geometry is final)
  r <- crop(r, clip_ext)
  
  # 2) crop mask too (cheap safety even if already cropped outside)
  m <- crop(mask_r, clip_ext)
  
  # 3) align mask to the cropped raster grid
  if (!compareGeom(r[[1]], m, stopOnError = FALSE)) {
    m <- resample(m, r[[1]], method = "near")
  }
  
  # 4) apply mask
  r <- mask(r, m)
  
  writeRaster(
    r, out_file, overwrite = TRUE,
    wopt = list(datatype = "FLT4S", gdal = c("COMPRESS=LZW","TILED=YES","BIGTIFF=IF_SAFER"))
  )
  invisible(out_file)
}
```


This is used for averaging 4 year average for the analysis. This function is written by Reeti Nykänen (reeti.nykanen@aalto.fi)
```{r}
# ------------------------------------------------------------------
# calculate 4 year average
# ------------------------------------------------------------------
calculate_4year_averages <- function(input_path, dataset_name, output_dir, years, periods, mask_r) {
  cat("averaging:", dataset_name, "\n")
  
  # open once (doesn't read all into memory)
  r_all <- rast(input_path)
  r_all <- crop(r_all, clip_extent)
  r_all <- ensure_year_names(r_all, years)
  
  # align mask to raster grid
  if (!compareGeom(r_all[[1]], mask_r, stopOnError = FALSE)) {
    mask2 <- resample(mask_r, r_all[[1]], method = "near")
  } else {
    mask2 <- mask_r
  }
  
  out_layers <- list()
  
  for (pnm in names(periods)) {
    sel_names <- paste0("y", periods[[pnm]])
    
    # read only 4 or 5 layers for this period
    r_sel <- r_all[[sel_names]]
    
    # special NA->0 inside mask BEFORE averaging
    if (dataset_name %in% na_to_zero_before_avg) {
      r_sel <- ifel(!is.na(mask2) & is.na(r_sel), 0, r_sel)
    }
    
    m <- mean(r_sel, na.rm = TRUE)
    names(m) <- pnm
    out_layers[[pnm]] <- m
    
    rm(r_sel, m)
    gc(FALSE)
  }
  
  final_stack <- rast(out_layers)
  output_file <- file.path(output_dir, paste0(dataset_name, "_1992-2020_4y.tif"))
  
  writeRaster(
    final_stack, output_file, overwrite = TRUE,
    wopt = list(datatype = "FLT4S", gdal = c("COMPRESS=LZW","TILED=YES","BIGTIFF=IF_SAFER"))
  )
  
  rm(r_all, final_stack, out_layers)
  gc(FALSE)
  
  output_file
}
```


This is used for cleaning the names of the files. This function is written by Reeti Nykänen (reeti.nykanen@aalto.fi)
```{r}
# ------------------------------------------------------------------
# function to clean the names
# ------------------------------------------------------------------
clean_name <- function(x) {
  x <- gsub("_1992-2020_4y\\.tif$", "", x)
  x <- gsub("\\.tif$", "", x)
  x
}
```


This is used for getting statistical quantile values from the datasets to help decide whether to fill in NA values before normalising or do a sanity check on the dataset. This function is written by Reeti Nykänen (reeti.nykanen@aalto.fi)
```{r}
# ------------------------------------------------------------------
# get the statistics of quantile values from the datasets before percentile clamping and normalising
# ------------------------------------------------------------------
get_stats_all_layers <- function(r, do_pct = FALSE) {
  v <- values(r)
  v <- v[is.finite(v)]
  if (length(v) == 0) return(list(p2.5 = NA, p97.5 = NA, mean = NA, sd = NA))
  
  if (do_pct) {
    p2.5  <- unname(quantile(v, 0.025, na.rm = TRUE))
    p97.5 <- unname(quantile(v, 0.975, na.rm = TRUE))
    v_cl  <- v[v >= p2.5 & v <= p97.5]
    if (length(v_cl) > 0) v <- v_cl
  } else {
    p2.5  <- NA_real_
    p97.5 <- NA_real_
  }
  
  m   <- mean(v, na.rm = TRUE)
  sdv <- sd(v, na.rm = TRUE)
  if (!is.finite(sdv) || sdv <= 0) sdv <- 1
  
  list(p2.5 = p2.5, p97.5 = p97.5, mean = m, sd = sdv)
}
```


This is used for normalising the dataset. This function is written by Reeti Nykänen (reeti.nykanen@aalto.fi)
```{r}
# ------------------------------------------------------------------
# normalising function: needs to run both clean_name function and get_stats_all_layers function first
# ------------------------------------------------------------------
normalize_one <- function(file_path) {
  dataset_name <- clean_name(basename(file_path))
  cat("\n--- normalizing:", dataset_name, "---\n")
  
  r <- rast(file_path)
  r <- crop(r, clip_extent)
  
  # enforce period names if missing
  if (nlyr(r) == 7) {
    nm <- names(r)
    if (is.null(nm) || any(nm == "") || !all(period_names %in% nm)) {
      names(r) <- period_names
    }
  }
  
  # align mask to raster grid
  if (!compareGeom(r[[1]], BOKUmask, stopOnError = FALSE)) {
    mask2 <- resample(BOKUmask, r[[1]], method = "near")
  } else {
    mask2 <- BOKUmask
  }
  
  # mask first
  r <- mask(r, mask2)
  
  # NA -> 0 only inside mask domain
  if (dataset_name %in% na_to_zero_inside_mask) {
    r <- ifel(!is.na(mask2) & is.na(r), 0, r)
  }
  
  # percentile clamp (optional)
  do_pct <- dataset_name %in% percentile_datasets
  stats  <- get_stats_all_layers(r, do_pct = do_pct)
  
  if (do_pct && is.finite(stats$p2.5) && is.finite(stats$p97.5)) {
    r <- clamp(r, lower = stats$p2.5, upper = stats$p97.5, values = TRUE)
  }
  
  # z-score normalize
  r_norm <- (r - stats$mean) / stats$sd
  r_norm <- clamp(r_norm, lower = -2, upper = 2, values = TRUE)
  
  # keep period names
  if (nlyr(r_norm) == 7) names(r_norm) <- period_names
  
  out_path <- file.path(norm_dir, paste0(dataset_name, "_norm.tif"))
  writeRaster(
    r_norm, out_path, overwrite = TRUE,
    wopt = list(datatype = "FLT4S", gdal = c("COMPRESS=LZW","TILED=YES","BIGTIFF=IF_SAFER"))
  )
  
  cat("wrote:", out_path, "\n")
  invisible(out_path)
}
```


This is used for noormalizing using min-max. This function is written by Reeti Nykänen (reeti.nykanen@aalto.fi)
```{r}
# -------------------------
# helper: min-max normalization
# -------------------------
minmaxnorm <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (!is.finite(rng[1]) || !is.finite(rng[2]) || rng[1] == rng[2]) {
    return(rep(0, length(x)))
  }
  (x - rng[1]) / (rng[2] - rng[1])
}
```


