---
title: "SOC_SE1_SE2"
output: html_document
date: "2026-05-21"
---

This code combined the Standard Element 1 (Data acquisition) and Standard Element 2 (Data processing) for the soil organic carbon (SOC) dataset from SoilGrids (Poggio et al. 2020)

Reference: Poggio, L., de Sousa, L. M., Batjes, N. H., Heuvelink, G. B. M., Kempen, B., Ribeiro, E., and Rossiter, D.: SoilGrids 2.0: producing soil information for the globe with quantified spatial uncertainty, SOIL, 7, 217–240, https://doi.org/10.5194/soil-7-217-2021, 2021.

Since SOC does not change much through out the year, we will use value from the averaged year for all the years from 1992 to 2020.

Establish the libraries, connections, and directories. We want layers from 0-30 cm (g/kg)
```{r}
# Fast approach using gdal_translate for 5 arc-minute resolution
library(gdalUtilities)
library(terra)

# 1. Force GDAL to retry if the server glitches (prevents Error 4/timeout issues)
Sys.setenv(GDAL_HTTP_MAX_RETRY = "10")
Sys.setenv(GDAL_HTTP_RETRY_DELAY = "3")

# --- Define remote sources and local destinations ---
voi <- 'soc'
layers <- c('soc_0-5cm_mean', 'soc_5-15cm_mean', 'soc_15-30cm_mean')
urls <- paste0("/vsicurl/https://files.isric.org/soilgrids/latest/data/", voi, '/', layers, '.vrt')

# Create the directory if it doesn't exist
out_dir <- "Documents/Doctoral/Research/dataset/SoilGrids/"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
local_files <- paste0(out_dir, layers, "_5arcmin.tif")
```

Download the dataset
```{r}
# 2. Download directly at 5 arc-minute resolution (4320 x 2160 pixels globally)
for (i in 1:length(urls)) {
  if (file.exists(local_files[i])) {
    file.remove(local_files[i])
  }
  
  # Run gdal_translate with outsize to instantly aggregate to 5 arc-minutes
  gdal_translate(
    src_dataset = urls[i],
    dst_dataset = local_files[i],
    of = "GTiff",
    outsize = c(4320, 2160), 
    r = "average",            # Ensures we average the SOC values rather than picking random pixels
    co = c("COMPRESS=DEFLATE", "TILED=YES")
  )
}
```


Since the datasets have three layers, we use weighted average to calculate SOC for 30cm depths
```{r}
# 3. Load local 5-arcmin rasters and compute the weighted average
soc5 <- rast(local_files[1])
soc15 <- rast(local_files[2])
soc30 <- rast(local_files[3])

# Vectorized calculation
soc0_30 <- (((5/30) * soc5 + (10/30) * soc15 + (15/30) * soc30)) / 10 # g/kg
```

Save the files
```{r}
# 4. Write the final computed 5-arcmin raster
writeRaster(
  soc0_30,
  paste0(out_dir, "soc_0-30cm_mean_5arcmin.tif"),
  overwrite = TRUE,
  wopt = list(
    filetype = "GTiff",
    gdal = c("COMPRESS=DEFLATE", "TILED=YES")
  )
)
```

Change the projection of the files into the final stack resolution
```{r}
# 5. Create raster reference and check out the raster files
r_ref <- rast(ncol=4320, nrows = 2160, crs   = "EPSG:4326")

out_dir <- "Documents/Doctoral/Research/dataset/SoilGrids/"

files_to_project <- list.files(out_dir, pattern = "_5arcmin\\.tif$", full.names = TRUE)

for (file in files_to_project) {
  
  # Load the raster
  current_rast <- rast(file)
  names(current_rast) <- tools::file_path_sans_ext(basename(file)) #To fix the name of the raster to align
  
  message("Projecting: ", basename(file))
  
  # Create a temporary filename
  temp_file <- paste0(file, ".tmp")
  
  # Project to the temporary file
  project(
    current_rast, 
    r_ref, 
    method = "bilinear", 
    filename = temp_file,
    overwrite = TRUE,
    wopt = list(
      filetype = "GTiff",
      gdal = c("COMPRESS=DEFLATE", "TILED=YES")
    )
  )
  
  # Replace the original file with the projected version
  file.remove(file)
  file.rename(temp_file, file)
}

message("All files successfully projected and replaced!")
```

Duplicating the layers into 29 layers from 1992 to 2020
```{r}
library(terra)

# Input single-layer raster
soc_single <- rast("soc_0-30cm_mean_5arcmin.tif")

# Duplicate to 29 layers (1992-2020)
soc_stack <- rast(replicate(29, soc_single, simplify = FALSE))

# Name layers y1992 ... y2020
names(soc_stack) <- paste0("y", 1992:2020)

# Save stack
writeRaster(
  soc_stack,
  "soc_0-30cm_mean_5arcmin_stack.tif",
  overwrite = TRUE
)
```