---
title: "Aridity_Index_SE2"
output: html_document
date: "2026-05-21"
---

Standard Element 2 refers to data processing. In this step we will harmonise the code to get the stack dataset ready to be used in the analysis process.

These code chunks are written by Ashvath Kunadi (ORCID ID: 0000-0002-3098-1247). It is meant to read the nc files for precipitation dataset from MSWEP (Wang et al. 2025) and potential evaporation from GLEAM (Miralles et al. 2025)

This chunk is to read the monthly precipitation data. You would need to run this code chunk first.
```{r}
prdir <- "D:/MSWEP_monthly/Monthly/"
pr_list <- list()
for(j in 1:length(years)){
  yr_fil <- list.files(prdir, pattern = paste0(years[j], '\\d{2}\\.nc'), full.names = TRUE)
  for(i in 1:length(yr_fil)){
    nc_name <- yr_fil[i]
    ncin <- nc_open(nc_name)
    time <- ncvar_get(ncin,"time")
    nt <- dim(time)
    lon <- ncvar_get(ncin,"lon")
    nlon <- dim(lon)
    lat <- ncvar_get(ncin,"lat")
    nlat <- dim(lat)
    mnth_rast <- ncvar_get(ncin, "precipitation")
    nc_close(ncin)
    if(i == 1){
      yr_rast <- mnth_rast
    }else{
      yr_rast <- yr_rast+mnth_rast
    }
  }
  pr_list[[j]] <- yr_rast
  # trial <- rast(vals = t(yr_rast), ncol=nlon, nrow=nlat, xmin= -179.95, xmax=179.95, ymin=-89.95, ymax=89.95)
  # writeRaster(trial, paste0(prdir, 'Precip_',years[j],'.tif'), overwrite=TRUE)
}
```


This second chunk is to read the potential evaporation dataset and then calculate the aridity index from the formula AI=P/PET
```{r}
prdir <- "D:/GLEAM_yearly/Ep/"
for(j in 1:length(years)){
  nc_name <- paste0(prdir, "Ep_",years[j],"_GLEAM_v4.2a_YR.nc")
  ncin <- nc_open(nc_name)
  time <- ncvar_get(ncin,"time")
  nt <- dim(time)
  lon <- ncvar_get(ncin,"lon")
  nlon <- dim(lon)
  lat <- ncvar_get(ncin,"lat")
  nlat <- dim(lat)
  yr_rast <- ncvar_get(ncin, "Ep")
  nc_close(ncin)
  
  ar_rast <- pr_list[[j]]/yr_rast #Calculation of aridity index. Needs pr_list from precipitation first
  trial <- rast(vals = t(ar_rast), ncol=nlon, nrow=nlat, xmin= -179.95, xmax=179.95, ymin=-89.95, ymax=89.95)
  writeRaster(trial, paste0(prdir, 'Ari_',years[j],'.tif'), overwrite=TRUE)
}
```


Chunks after this is written by me ("Best" Bhattarabhop Viriyaroj)

First we read the file and convert it into 5 arcminute
```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
library(terra)
library(tools)

# 1. Define the target 5-arcminute reference raster
# This creates an empty raster with the desired dimensions and projection.
r_ref <- rast(ncol = 4320, nrow = 2160, crs = "EPSG:4326")

# 2. Set paths for input and output files
# Assumes your aridity files are in a directory named "Aridity"
# and you want to save the new files to "Aridity_5arcmin".
in_dir <- "Aridity"
out_dir <- "Aridity_5arcmin"

# Create the output directory if it doesn't exist
if (!dir.exists(out_dir)) {
  dir.create(out_dir)
}

# Get a list of all aridity files to process
aridity_files <- list.files(
  path = in_dir,
  pattern = "\\.tif$",  # Matches files like Ari_1992.tif
  full.names = TRUE
)

# 3. Loop through, resample, and save each raster
for (file_path in aridity_files) {
  # Read the original 6-arcminute raster
  r_original <- rast(file_path)
  
  # Resample the raster to match the 5-arcminute reference raster
  # 'bilinear' is a good method for continuous data like an aridity index.
  r_resampled <- resample(r_original, r_ref, method = "bilinear")
  
  # Define the output filename and path
  base_name <- file_path_sans_ext(basename(file_path))
  
  # Create the new filename with the "_5arcmin" suffix
  out_filename <- paste0(base_name, "_5arcmin.tif")
  out_path <- file.path(out_dir, out_filename)
  
  # Save the resampled raster
  writeRaster(r_resampled, out_path, overwrite = TRUE)
  
  # Print a message to track progress
  message("Resampled and saved: ", out_filename)
}

message("Resampling complete for all files.")
```

Since the dataset is extremely skewed and for aridity index, the lower values of the datasets is to be focused more since it shows how arid certain areas are. We are less interested in non-arid area since they perform similarly (refer to Zomer et al. 2022)

Thus, we decided transform the dataset using logarithm.
```{r}
in_dir <- "Aridity_5arcmin"
out_dir <- "Aridity_5arcmin/log_transform"

# Create the output directory if it doesn't exist
if (!dir.exists(out_dir)) {
  dir.create(out_dir)
}

# 2. Get a list of all 5-arcminute aridity files
files_to_transform <- list.files(
  path = in_dir,
  pattern = "_5arcmin\\.tif$",
  full.names = TRUE
)

# 3. Loop through, transform, and save each raster
for (file_path in files_to_transform) {
  # Read the original 5-arcminute raster
  r_original <- rast(file_path)
  
  # Apply the log transformation
  # Note: log() of values <= 0 will produce -Inf or NaN.
  # This is usually fine, as they will be treated as NA by default.
  r_log <- log(r_original)
  
  base_name <- file_path_sans_ext(basename(file_path))
  
  # Define the output path (same filename, new directory)
  out_filename <- paste0(base_name, "_log.tif")
  out_path <- file.path(out_dir, out_filename)
  
  # Save the log-transformed raster
  writeRaster(r_log, out_path, overwrite = TRUE)
  
  message("Log-transformed and saved: ", out_filename)
}

message("Log transformation complete for all files.")
```

We then stack the files
```{r}
library(terra)

# 1. Set the directory containing the log-transformed files
in_dir <- "Aridity_5arcmin/log_transform"

# 2. Get a sorted list of the files to stack
files_to_stack <- list.files(
  path = in_dir,
  pattern = "_log\\.tif$",
  full.names = TRUE
) |> sort()

# 3. Create the SpatRaster stack
aridity_log_stack <- rast(files_to_stack)

# 4. Create and assign the new layer names
new_names <- paste0("y", 1992:2020)
names(aridity_log_stack) <- new_names

# 5. Define the output path and save the raster stack
out_path <- file.path("Aridity_5arcmin", "aridity_log_stack_1992-2020.tif")
writeRaster(aridity_log_stack, out_path, overwrite = TRUE)

# Print a confirmation message
message("Raster stack saved to: ", out_path)

# Display the new stack's information
aridity_log_stack
plot(aridity_log_stack)
```


