---
title: "GDD_SE2"
output: html_document
date: "2026-05-22"
---

This document shows how we can get the calculate the growing degree days dataset and harmonise it into 5 arcminute from 1992 to 2020

Call the necessary libraries and set the paths
```{r}
library(terra)
library(lubridate)
library(dplyr)
library(tidyr)
library(sf)
library(readxl)
library(terra)
library(dplyr)
library(readr)

#Call the paths and set working directory
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

# Paths (adjust to yours)
path_tmax_5 <- "/Volumes/MainEx/BestMemory/Research/dataset/MSWX/MSWX_daily/rasterfiles/tmax/5arcmin/"
path_tmin_5 <- "/Volumes/MainEx/BestMemory/Research/dataset/MSWX/MSWX_daily/rasterfiles/tmin/5arcmin/"
path_cropmask <- "CropMask/5arcmin/"
path_gdd <- "GDD_rasterfiles/"  # Output path (adjust if different)
```

Call the necessary functions
```{r}
# Function to extract masked values as data.frame (only masked cells, columns per day)
myFun_toVector <- function(rastIn, mask) {
  mask1 <- mask[[1]]
  cells <- which(values(mask1) == 1)
  if (length(cells) == 0) return(data.frame())
  vals <- terra::extract(rastIn, cells)  # Returns data.frame with ID + layer cols
  vals <- vals[, -1, drop = FALSE]       # Drop ID column
  return(vals)
}

# Function to compute annual GDD from masked tmax/tmin data.frames
compute_gdd_df <- function(tmax_df, tmin_df, Tbase, Tcutoff) {
  # Vectorized per cell/day: cap tx, floor tn, avg, subtract base, max(0), sum over days
  tx_cap <- pmin(tmax_df, Tcutoff)
  tn_cap <- pmax(tmin_df, Tbase)
  tavg <- (tx_cap + tn_cap) / 2
  gdd_daily <- pmax(tavg - Tbase, 0)
  gdd_annual <- rowSums(gdd_daily, na.rm = TRUE)  # Sum over days per cell
  return(gdd_annual)
}
```

First we need to calculate the fraction of HANPPharv first in order to do weighted average.
```{r}
library(terra)
library(stringr)

# Define paths
harvest_dir <- "../LUIcube_BOKU/HANPPharv/5arcminute"
yearly_sum_dir <- file.path(harvest_dir, "yearly_sum")
fraction_out_dir <- "CropFraction/"

# Create output directory if it doesn't exist
if (!dir.exists(fraction_out_dir)) {
  dir.create(fraction_out_dir, recursive = TRUE)
}

# Get list of all individual crop harvest files
all_harvest_files <- list.files(
  path = harvest_dir,
  pattern = "\\.tif$",
  full.names = TRUE,
  recursive = FALSE # Only files in the main directory
)

# Filter out fodder crops and any files from subdirectories
all_harvest_files <- all_harvest_files[!grepl("CL-FODD", all_harvest_files)]
all_harvest_files <- all_harvest_files[!grepl("yearly_sum", all_harvest_files)]


# Group files by year
years <- str_extract(basename(all_harvest_files), "^\\d{4}")
files_by_year <- split(all_harvest_files, years)

# Process each year
for (yr in names(files_by_year)) {
  message("Processing year: ", yr)
  
  # Load the total harvest raster for the year
  total_harvest_file <- file.path(yearly_sum_dir, paste0("HANPPharv_tC_", yr, "_5arcminute.tif"))
  if (!file.exists(total_harvest_file)) {
    warning("Skipping year ", yr, ": yearly sum file not found.")
    next
  }
  total_harvest <- rast(total_harvest_file)
  
  # To avoid division by zero, set cells with zero total harvest to NA
  # This ensures that fractions are only calculated where there is harvest
  total_harvest[total_harvest == 0] <- NA
  
  # Get the list of crop files for the current year
  crop_files_yr <- files_by_year[[yr]]
  
  # Process each crop file for the year
  for (crop_file in crop_files_yr) {
    crop_name_match <- str_match(basename(crop_file), "_tC_CL-([A-Z0-9]+)_")
    if (is.na(crop_name_match[1,2])) {
        next
    }
    crop_code <- crop_name_match[1,2]
    
    message("  Calculating fraction for: ", crop_code)
    
    # Define output file path
    out_file <- file.path(fraction_out_dir, paste0("fraction_", crop_code, "_", yr, ".tif"))
    
    if (file.exists(out_file)) {
        message("    - SKIPPED (output file already exists)")
        next
    }

    # Load individual crop harvest raster
    crop_harvest <- rast(crop_file)
    
    # Calculate fraction
    crop_fraction <- crop_harvest / total_harvest
    
    # Write the fraction raster to the output directory
    writeRaster(crop_fraction, out_file, overwrite = TRUE, datatype = "FLT4S")
    message("    - Saved: ", out_file)
  }
}

message("\nCrop fraction calculation complete.")
```


Main loop over crops and years

There is a main limitation here. From the file LUIcube_cropBaseCutoffTemperature_v2.xlsx, you can see the calculation for the growth range for each crop class. However, based on values from Peredes et al. 2025 not all of the crops that appeared in the crop class have growing degree days range. This means that the growing degree range has to be averaged for those crop classes that have multiple crops. A better practice would be to do a weighted average, however, some of the crop groups have less than half of the available information. Thus, we will use only normal average. Check out the excel file if you want to see the process.

```{r}
# --- Main Loop ---

# Read crop parameters
crops_tbl <- read_excel("LUIcube_cropBaseCutoffTemperature_v2.xlsx")

# Loop over each crop
for (crop_idx in seq_len(nrow(crops_tbl))) {
  crop_name <- crops_tbl$crop_code[crop_idx]
  Tbase <- crops_tbl$Tbase[crop_idx]
  Tcutoff <- crops_tbl$Tcutoff[crop_idx]
  
  message("\n\n======= PROCESSING CROP: ", crop_name, " (Tbase=", Tbase, ", Tcutoff=", Tcutoff, ") =======")
  
  # Loop over years
  for (yr in 1992:2020) {
    out_file <- file.path(path_gdd, paste0("GDD_weighted_", crop_name, "_", yr, ".tif"))
    
    if (file.exists(out_file)) {
      message("  Year ", yr, " - SKIPPED (output file already exists)")
      next
    }
    
    message("  Year ", yr, "...")
    
    # Load tmax and tmin rasters
    tmax_file <- file.path(path_tmax_5, paste0("tmax_", yr, ".tif"))
    tmin_file <- file.path(path_tmin_5, paste0("tmin_", yr, ".tif"))
    if (!file.exists(tmax_file) || !file.exists(tmin_file)) {
      warning("Skipping ", yr, ": tmax/tmin files missing")
      next
    }
    tmax <- rast(tmax_file)
    tmin <- rast(tmin_file)
    
    # Find mask file (used to define which cells to process)
    mask_file <- list.files(path_cropmask, 
                            pattern = paste0("^", yr, ".*", crop_name, ".*_mask\\.tif$"), 
                            full.names = TRUE)[1]
    if (is.na(mask_file)) {
      warning("  Skipping ", yr, ": mask missing for ", crop_name)
      next
    }
    mask <- rast(mask_file)
    
    # MODIFIED: Find and load the corresponding crop fraction file
    fraction_file <- file.path(path_cropfraction, paste0("fraction_", crop_name, "_", yr, ".tif"))
    if (!file.exists(fraction_file)) {
        warning("  Skipping ", yr, ": fraction file missing for ", crop_name)
        next
    }
    crop_fraction <- rast(fraction_file)

    # Extract data for masked cells
    tmax_df <- myFun_toVector(tmax, mask)
    tmin_df <- myFun_toVector(tmin, mask)
    # MODIFIED: Extract fraction values for the same masked cells
    crop_fraction_df <- myFun_toVector(crop_fraction, mask)

    if (nrow(tmax_df) == 0) {
      warning("  Skipping ", yr, ": no masked cells")
      next
    }
    
    # MODIFIED: Compute weighted annual GDD per cell
    gdd_vals <- compute_gdd_df(tmax_df, tmin_df, crop_fraction_df, Tbase, Tcutoff)
    
    # Project back to raster
    gdd_raster <- mask
    gdd_raster[] <- NA
    cells <- which(values(mask) == 1)
    gdd_raster[cells] <- gdd_vals
    
    # Write output
    writeRaster(gdd_raster, out_file, overwrite = TRUE, datatype = "FLT4S")
    message("    ✓ Saved: ", out_file)
    
    # Clean up
    rm(tmax, tmin, mask, crop_fraction, tmax_df, tmin_df, crop_fraction_df, gdd_vals, gdd_raster)
    gc()
  }
}

message("\n\n======= ALL WEIGHTED GDD PROCESSING COMPLETE =======\n")
```

Stack the file into yearly sum and stack
```{r}
library(terra)

in_dir <- "GDD_rasterfiles"
out_dir <- file.path(in_dir, "yearly_sum")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# Files like: GDD_banana_1992.tif, GDD_barley_1992.tif, ...
all_files <- list.files(
  path = in_dir,
  pattern = "^GDD_.*_(19[0-9]{2}|20[0-9]{2})\\.tif$",
  full.names = TRUE
)

# Extract year from filename
years <- as.integer(sub(".*_(19[0-9]{2}|20[0-9]{2})\\.tif$", "\\1", basename(all_files)))
year_levels <- sort(unique(years))

out_files <- lapply(year_levels, function(yr) {
  f_yr <- all_files[years == yr]
  r_yr <- rast(f_yr)                    # multilayer (all crops for that year)
  r_sum <- app(r_yr, sum, na.rm = TRUE) # one layer: sum across crops
  names(r_sum) <- paste0("y", yr)
  
  out_file <- file.path(out_dir, paste0("GDD_", yr, ".tif"))
  writeRaster(r_sum, out_file, overwrite = TRUE)
  out_file
})

unlist(out_files)

#Then stack the files
r_gdd_stack <- function_raster_stack(in_directory = "GDD_rasterfiles/yearly_sum")
```